Preferences for mercantilization: the role of intergenerational social mobility and beliefs in meritocracy in Chile

### 2025-06-27 advances/informe2.qmd
Up informe 3

---
title: "Preferences for mercantilization: the role of intergenerational social mobility and beliefs in meritocracy in Chile"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
    code-link: true
    title-block-banner: true
    theme:
      - cosmo
      - pretty.scss
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "../input/bib/tesis-magister.bib"
csl: "../input/bib/apa6.csl"
author:
  - name: Andreas Laffert Tamayo
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 1
affiliations: 
  - id: 1
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
citeproc: true
link-citations: true
colorlinks: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
fig-height: 4
fig-width: 7.5
---

@@

What is the legitimate extent of market inequality in the eyes of the public? Since the early 1980s, many countries have experienced a widespread retreat from universal welfare programs and a shift toward the privatization and commodification of public goods, welfare policies, and social services [@gingrich_making_2011; @streeck_how_2016]. In Latin America, as elsewhere, neoliberal reforms reshaped welfare-state institutions by extending market logic into domains of social reproduction that were traditionally governed by the state [@arrizabalo_milagro_1995; @ferre_welfare_2023]. This transformation reduced the role of public provision and increased the presence of private actors in core social services [@harvey_breve_2015]. Echoing Polanyi's [-@polanyi_great_1975] insight that markets constitute a distinct moral order, the institutional diffusion of market rules has fostered a corresponding moral economy: a constellation of norms and values concerning fair allocation, embedded in institutions and shaping individual subjectivities [@mau_inequality_2015; @svallforsMoralEconomyClass2006a]. Within this framework, a growing body of research examines the extent to which, and the mechanisms by which, citizens consider it fair that the allocation of services like health care, pensions, and education be governed by market-based criteria---a phenomenon known as *market justice preferences* [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015]. Understanding these preferences is crucial, as they contribute to legitimizing economic inequality by framing it as the fair result of individual responsibility and limited state intervention [@mau_inequality_2015].

@@

Although redistribution typically refers to the state’s capacity to reallocate resources from the advantaged to the vulnerable, market institutions also play a central role in distributing socially valuable goods and rewards [@koos_moral_2019; @lindh_bringing_2023]. As Polanyi [-@polanyi_great_1975] observed, economic integration in capitalist societies is primarily organized through market exchange, governed by a self-regulating price system embedded in institutional frameworks. These institutions are not mere aggregates of individual behavior, but social realities endowed with rules, mechanisms, and normative meanings that shape everyday thinking [@koos_moral_2019; @immergut_theoretical_1998]. In this sense, the economic order is mirrored in a moral economy: collectively shared norms and beliefs about justice in distribution, embedded and reinforced through institutions [@mau_inequality_2015; @svallfors_political_2007]. While most research from this perspective has focused on welfare institutions, recent scholarship has brought the market back into focus as a site of distributive justice beliefs and institutional responses to inequality [@immergut_theoretical_1998]. In many countries, privatization and commodification have expanded market logic into core areas of social reproduction, such as healthcare, education, pensions, and social security, deepening inequality in access to these services [@gingrich_making_2011; @ferre_welfare_2023]. Yet public support for market-based welfare provision has grown, even in traditional welfare states, particularly among higher-income groups who view private alternatives as more efficient or higher in quality [@busemeyer_welfare_2020]. This shift calls for broader inquiry into how institutions like markets structure access to resources and legitimize inequalities [@lindh_public_2015; @mau_inequality_2015].

The legitimacy of market-based inequalities is closely tied to beliefs about distributive justice grounded in market principles. Beyond stratification structures and socioeconomic conditions, the study of inequality also examines the beliefs individuals hold about the causes of social inequality, the normative principles that underpin those beliefs, the factors that shape them, and their consequences for attitudes and behavior [@kluegel_beliefs_1981]. In this regard, empirical research on distributive justice focuses on individuals’ conceptions of how goods and rewards _should_ be distributed in society [@jasso_distributive_2016]. This line of inquiry allows for examining the extent to which economic inequality is perceived as just, or in a certain sense, legitimate [@castillo_legitimacy_2011]. One prominent strand investigates the legitimacy of wage inequality, particularly salary differences across occupations [@castillo_legitimacy_2011; @jasso_gender_1999; @jasso_justice_1978]. Another examines how individuals justify market-generated inequalities in access to core social services such as healthcare, education, and pensions. Here, legitimacy stems from the belief that access to these goods should follow market-based criteria [@castillo_socialization_2024;  @lindh_public_2015]. In such views, these services are treated as legitimate commodities; goods that can be traded, priced, and evaluated through market logic [@busemeyer_welfare_2020].

@@

The empirical study of market justice preferences has relied on diverse strategies to operationalize how individuals evaluate inequalities produced by market allocation. One common approach assesses whether people consider it fair that access to key social services—such as healthcare, education, or pensions—depends on income. This strategy builds on foundational work by Kluegel and Smith [-@kluegel_beliefs_1981], who explored the normative justifications behind support for economic inequality. More recent studies have expanded this logic beyond income to include market-mediated access to welfare goods. For instance, Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al. [-@vondemknesebeck_are_2016] examine whether respondents believe it is just that those with higher incomes receive better healthcare. Similarly, Lee and Stacey [-@lee_fairness_2023] apply this approach to educational opportunities. These studies typically use targeted survey items that ask respondents to evaluate specific distributive scenarios, allowing for comparative analyses across countries or welfare regimes. Complementing these efforts, more recent research has introduced composite indicators designed to capture broader orientations toward market-based allocation. For example, Castillo et al. [-@castillo_socialization_2024] propose a single-item measure that summarizes individual support for income-based access across multiple domains—health, education, and pensions—within the Chilean context. Together, these instruments aim to capture to what extent individuals perceive market-generated inequality as justified.

@@

Research on intergenerational social mobility has increasingly examined its effects on attitudes toward economic inequality. A key area of inquiry focuses on how upward and downward mobility influence support for redistribution, though findings remain mixed. Alesina et al. [-@alesina_intergenerational_2018] show that individuals with pessimistic expectations about their mobility—particularly those anticipating downward movement—are more likely to support generous redistributive policies. Similarly, Ares [-@ares_changing_2020] finds that upwardly mobile individuals tend to be less supportive of state-led redistribution compared to those who have experienced downward mobility. Comparative studies by Schmidt [-@schmidt_experience_2011] and Gugushvili [@gugushvili_subjective_2017] likewise report that subjective upward mobility is associated with weaker preferences for redistribution, while downward mobility strengthens redistributive support. However, recent causal evidence from Breen and Ermisch [-@breen_effects_2024] suggests the opposite: upward mobility may increase redistributive preferences, while downward mobility may reduce them.

@@

These findings suggest that causal attribution processes mediate the relationship between mobility experiences and justice-related attitudes. Accordingly, this study proposes the following hypothesis:

>>>>

---
title: "Preferences for mercantilization: the role of intergenerational social mobility and beliefs in meritocracy in Chile"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
    code-link: true
    title-block-banner: true
    theme:
      - pretty.scss
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "../input/bib/tesis-magister.bib"
csl: "../input/bib/apa6.csl"
author:
  - name: Andreas Laffert Tamayo
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 1
affiliations: 
  - id: 1
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
link-citations: true
linestretch: 1.5       
mainfont: Times New Roman
fontsize: 13pt          
fig-height: 4
fig-width: 7.5
---

@@

What is the legitimate extent of market inequality in the eyes of the public? Since the early 1980s, many countries have experienced a widespread retreat from universal welfare programs and a shift toward the privatization and commodification of public goods, welfare policies, and social services [@gingrich_making_2011; @streeck_how_2016]. In Latin America, as elsewhere, neoliberal reforms reshaped welfare-state institutions by extending market logic into domains of social reproduction that were traditionally governed by the state [@ferre_welfare_2023]. This transformation reduced the role of public provision and increased the presence of private actors in core social services [@harvey_breve_2015]. Echoing Polanyi's [-@polanyi_great_1975] insight that markets constitute a distinct moral order, the institutional diffusion of market rules has fostered a corresponding moral economy: a constellation of norms and values concerning fair allocation, embedded in institutions and shaping individual subjectivities [@mau_inequality_2015; @svallforsMoralEconomyClass2006a]. Within this framework, a growing body of research examines the extent to which, and the mechanisms by which, citizens consider it fair that the allocation of services like health care, pensions, and education be governed by market-based criteria---a phenomenon known as *market justice preferences* [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015]. Understanding these preferences is crucial, as they contribute to legitimizing economic inequality by framing it as the fair result of individual responsibility and limited state intervention [@mau_inequality_2015].

@@

Beyond state’s capacity to reallocate resources from the advantaged to the vulnerable, market institutions also play a central role in distributing socially valuable goods and rewards [@koos_moral_2019; @lindh_bringing_2023]. As Polanyi [-@polanyi_great_1975] observed, economic integration in capitalist societies is primarily organized through market exchange, governed by a self-regulating price system embedded in institutional frameworks. These institutions are not mere aggregates of individual behavior, but social realities endowed with rules, mechanisms, and normative meanings that shape everyday thinking [@koos_moral_2019; @immergut_theoretical_1998]. In this sense, the economic order is mirrored in a moral economy: collectively shared norms and beliefs about justice in distribution, embedded and reinforced through institutions [@mau_inequality_2015; @svallfors_political_2007]. While most research from this perspective has focused on welfare institutions, recent scholarship has brought the market back into focus as a site of distributive justice beliefs and institutional responses to inequality [@lindh_bringing_2023]. In many countries, privatization and commodification have expanded market logic into core areas of social reproduction, such as healthcare, education, pensions, and social security, deepening inequality in access to these services [@gingrich_making_2011; @ferre_welfare_2023]. Yet public support for market-based welfare provision has grown, even in traditional welfare states, particularly among higher-income groups who view private alternatives as more efficient or higher in quality [@busemeyer_welfare_2020]. This shift calls for broader inquiry into how institutions like markets structure access to resources and legitimize inequalities [@satz_por_2019; @mau_inequality_2015].

The legitimacy of market-based inequalities is closely tied to beliefs about distributive justice grounded in market principles. Beyond stratification structures and socioeconomic conditions, the study of inequality also focuses on individuals’ beliefs about the origins of inequality, the normative frameworks that sustain those beliefs, the mechanisms that shape them, and their implications for attitudes and behavior [@kluegel_beliefs_1981]. In this regard, empirical research on distributive justice focuses on individuals’ conceptions of how goods and rewards _should_ be distributed in society [@jasso_distributive_2016]. This line of inquiry allows for examining the extent to which economic inequality is perceived as just, or in a certain sense, legitimate [@castillo_legitimacy_2011]. Most of the research on this field explores the legitimacy of wage inequality, specially salary gaps between jobs at opposite ends of the occupational hierarchy [@castillo_legitimacy_2011; @jasso_gender_1999; @jasso_justice_1978]. Recently, others examines how individuals justify market-generated inequalities in access to core social services such as healthcare, education, pensions or social security. Here, legitimacy stems from the belief that access to these goods should follow market-based criteria [@castillo_socialization_2024;  @lindh_public_2015]. In such views, these services are treated as legitimate commodities; goods that can be traded, priced, and evaluated through market logic [@busemeyer_welfare_2020].

@@

The empirical study of market justice preferences has employed various strategies to capture how individuals assess inequalities arising from market allocation. A common approach evaluates whether people find it fair that access to essential social services—such as healthcare, education, or pensions—depends on income. This builds on foundational work by Kluegel and Smith [-@kluegel_legitimation_1999], who examined normative justifications for capitalist systems. Subsequent studies have extended this logic to a wider set of welfare goods. For example, Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al. [-@vondemknesebeck_are_2016] explore whether respondents believe it is just that higher-income individuals receive better healthcare. Similarly, Lee and Stacey [-@lee_fairness_2023] apply this framework to education. Complementing these efforts, more recent research has introduced composite indicators to capture broader orientations toward market-based allocation. Lindh [-@lindh_public_2015], for instance, constructs an index averaging support for income-based access to healthcare and education in comparative perspective, while Castillo et al. [-@castillo_socialization_2024] propose a single-item measure covering health, education, and pensions in Chile. These instruments seek to gauge to what extent individuals view market-generated inequalities as legitimate. Taken together, they capture two core dimensions of market distribution: the role of economic resources as a key determinant of outcomes, and the framing of services as tradable commodities that can be bought and sold according to ability to pay [-@lindh_public_2015].

@@

Research on intergenerational social mobility has increasingly examined its effects on attitudes toward economic inequality. A key area of inquiry focuses on how upward and downward mobility influence support for redistribution, though findings remain mixed. Alesina et al. [-@alesina_intergenerational_2018] show that individuals with pessimistic expectations about their mobility—particularly those anticipating downward movement—are more likely to support generous redistributive policies. Similarly, Ares [-@ares_changing_2020] finds that upwardly mobile individuals tend to be less supportive of state-led redistribution compared to those who have experienced downward mobility. Comparative studies by Schmidt [-@schmidt_experience_2011] and Gugushvili [-@gugushvili_subjective_2017] likewise report that subjective upward mobility is associated with weaker preferences for redistribution, while downward mobility strengthens redistributive support. However, recent causal evidence from Breen and Ermisch [-@breen_effects_2024] suggests the opposite: upward mobility may increase redistributive preferences, while downward mobility may reduce them.

@@

Consequently, considering this theoretical and empirical background, the first hypothesis of this research is that:

### 2025-06-27 advances/informe2.tex
Up informe 3

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{article}

@@

\setstretch{1.15}
\section{Introduction}\label{introduction}

What is the legitimate extent of market inequality in the eyes of the
public? Since the early 1980s, many countries have experienced a
widespread retreat from universal welfare programs and a shift toward
the privatization and commodification of public goods, welfare policies,
and social services (\citeproc{ref-gingrich_making_2011}{Gingrich,
2011}; \citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin
America, as elsewhere, neoliberal reforms reshaped welfare-state
institutions by extending market logic into domains of social
reproduction that were traditionally governed by the state
(\citeproc{ref-arrizabalo_milagro_1995}{Arrizabalo, 1995};
\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). This transformation
reduced the role of public provision and increased the presence of
private actors in core social services
(\citeproc{ref-harvey_breve_2015}{Harvey, 2015}). Echoing Polanyi's
(\citeproc{ref-polanyi_great_1975}{1975}) insight that markets
constitute a distinct moral order, the institutional diffusion of market
rules has fostered a corresponding moral economy: a constellation of
norms and values concerning fair allocation, embedded in institutions
and shaping individual subjectivities
(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). Within
this framework, a growing body of research examines the extent to which,
and the mechanisms by which, citizens consider it fair that the
allocation of services like health care, pensions, and education be
governed by market-based criteria---a phenomenon known as \emph{market
justice preferences} (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
2014}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
2024}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
Understanding these preferences is crucial, as they contribute to
legitimizing economic inequality by framing it as the fair result of
individual responsibility and limited state intervention
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

@@

Although redistribution typically refers to the state's capacity to
reallocate resources from the advantaged to the vulnerable, market
institutions also play a central role in distributing socially valuable
goods and rewards (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh,
2019}; \citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). As
Polanyi (\citeproc{ref-polanyi_great_1975}{1975}) observed, economic
integration in capitalist societies is primarily organized through
market exchange, governed by a self-regulating price system embedded in
institutional frameworks. These institutions are not mere aggregates of
individual behavior, but social realities endowed with rules,
mechanisms, and normative meanings that shape everyday thinking
(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019}). In this sense,
the economic order is mirrored in a moral economy: collectively shared
norms and beliefs about justice in distribution, embedded and reinforced
through institutions (\citeproc{ref-mau_inequality_2015}{Mau, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). While most
research from this perspective has focused on welfare institutions,
recent scholarship has brought the market back into focus as a site of
distributive justice beliefs and institutional responses to inequality
(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998}). In many
countries, privatization and commodification have expanded market logic
into core areas of social reproduction, such as healthcare, education,
pensions, and social security, deepening inequality in access to these
services (\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
\citeproc{ref-gingrich_making_2011}{Gingrich, 2011}). Yet public support
for market-based welfare provision has grown, even in traditional
welfare states, particularly among higher-income groups who view private
alternatives as more efficient or higher in quality
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).
This shift calls for broader inquiry into how institutions like markets
structure access to resources and legitimize inequalities
(\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

The legitimacy of market-based inequalities is closely tied to beliefs
about distributive justice grounded in market principles. Beyond
stratification structures and socioeconomic conditions, the study of
inequality also examines the beliefs individuals hold about the causes
of social inequality, the normative principles that underpin those
beliefs, the factors that shape them, and their consequences for
attitudes and behavior (\citeproc{ref-kluegel_beliefs_1981}{Kluegel \&
Smith, 1981}). In this regard, empirical research on distributive
justice focuses on individuals' conceptions of how goods and rewards
\emph{should} be distributed in society
(\citeproc{ref-jasso_distributive_2016}{Guillermina Jasso et al.,
2016}). This line of inquiry allows for examining the extent to which
economic inequality is perceived as just, or in a certain sense,
legitimate (\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011}).
One prominent strand investigates the legitimacy of wage inequality,
particularly salary differences across occupations
(\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011};
\citeproc{ref-jasso_justice_1978}{Guillermina Jasso, 1978};
\citeproc{ref-jasso_gender_1999}{G. Jasso \& Wegener, 1999}). Another
examines how individuals justify market-generated inequalities in access
to core social services such as healthcare, education, and pensions.
Here, legitimacy stems from the belief that access to these goods should
follow market-based criteria
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such views, these
services are treated as legitimate commodities; goods that can be
traded, priced, and evaluated through market logic
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).

@@

The empirical study of market justice preferences has relied on diverse
strategies to operationalize how individuals evaluate inequalities
produced by market allocation. One common approach assesses whether
people consider it fair that access to key social services---such as
healthcare, education, or pensions---depends on income. This strategy
builds on foundational work by Kluegel and Smith
(\citeproc{ref-kluegel_beliefs_1981}{1981}), who explored the normative
justifications behind support for economic inequality. More recent
studies have expanded this logic beyond income to include
market-mediated access to welfare goods. For instance, Immergut and
Schneider (\citeproc{ref-immergut_it_2020}{2020}) and von dem Knesebeck
et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016}) examine whether
respondents believe it is just that those with higher incomes receive
better healthcare. Similarly, Lee and Stacey
(\citeproc{ref-lee_fairness_2023}{2023}) apply this approach to
educational opportunities. These studies typically use targeted survey
items that ask respondents to evaluate specific distributive scenarios,
allowing for comparative analyses across countries or welfare regimes.
Complementing these efforts, more recent research has introduced
composite indicators designed to capture broader orientations toward
market-based allocation. For example, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) propose a single-item
measure that summarizes individual support for income-based access
across multiple domains---health, education, and pensions---within the
Chilean context. Together, these instruments aim to capture to what
extent individuals perceive market-generated inequality as justified.

@@

Research on intergenerational social mobility has increasingly examined
its effects on attitudes toward economic inequality. A key area of
inquiry focuses on how upward and downward mobility influence support
for redistribution, though findings remain mixed. Alesina et al.
(\citeproc{ref-alesina_intergenerational_2018}{2018}) show that
individuals with pessimistic expectations about their
mobility---particularly those anticipating downward movement---are more
likely to support generous redistributive policies. Similarly, Ares
(\citeproc{ref-ares_changing_2020}{2020}) finds that upwardly mobile
individuals tend to be less supportive of state-led redistribution
compared to those who have experienced downward mobility. Comparative
studies by Schmidt (\citeproc{ref-schmidt_experience_2011}{2011}) and
Gugushvili (\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017})
likewise report that subjective upward mobility is associated with
weaker preferences for redistribution, while downward mobility
strengthens redistributive support. However, recent causal evidence from
Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024}) suggests the
opposite: upward mobility may increase redistributive preferences, while
downward mobility may reduce them.

@@

\bibitem[\citeproctext]{ref-arrizabalo_milagro_1995}
Arrizabalo, X. (1995). \emph{{Milagro o quimera: la econom{í}a chilena
durante la dictadura}}. Libros de la Catarata.

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  13pt,
]{article}

@@

\setstretch{1.5}
\section{Introduction}\label{introduction}

What is the legitimate extent of market inequality in the eyes of the
public? Since the early 1980s, many countries have experienced a
widespread retreat from universal welfare programs and a shift toward
the privatization and commodification of public goods, welfare policies,
and social services (\citeproc{ref-gingrich_making_2011}{Gingrich,
2011}; \citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin
America, as elsewhere, neoliberal reforms reshaped welfare-state
institutions by extending market logic into domains of social
reproduction that were traditionally governed by the state
(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). This transformation
reduced the role of public provision and increased the presence of
private actors in core social services
(\citeproc{ref-harvey_breve_2015}{Harvey, 2015}). Echoing Polanyi's
(\citeproc{ref-polanyi_great_1975}{1975}) insight that markets
constitute a distinct moral order, the institutional diffusion of market
rules has fostered a corresponding moral economy: a constellation of
norms and values concerning fair allocation, embedded in institutions
and shaping individual subjectivities
(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). Within
this framework, a growing body of research examines the extent to which,
and the mechanisms by which, citizens consider it fair that the
allocation of services like health care, pensions, and education be
governed by market-based criteria---a phenomenon known as \emph{market
justice preferences} (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
2014}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
2024}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
Understanding these preferences is crucial, as they contribute to
legitimizing economic inequality by framing it as the fair result of
individual responsibility and limited state intervention
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

@@

Beyond state's capacity to reallocate resources from the advantaged to
the vulnerable, market institutions also play a central role in
distributing socially valuable goods and rewards
(\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). As Polanyi
(\citeproc{ref-polanyi_great_1975}{1975}) observed, economic integration
in capitalist societies is primarily organized through market exchange,
governed by a self-regulating price system embedded in institutional
frameworks. These institutions are not mere aggregates of individual
behavior, but social realities endowed with rules, mechanisms, and
normative meanings that shape everyday thinking
(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019}). In this sense,
the economic order is mirrored in a moral economy: collectively shared
norms and beliefs about justice in distribution, embedded and reinforced
through institutions (\citeproc{ref-mau_inequality_2015}{Mau, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). While most
research from this perspective has focused on welfare institutions,
recent scholarship has brought the market back into focus as a site of
distributive justice beliefs and institutional responses to inequality
(\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). In many
countries, privatization and commodification have expanded market logic
into core areas of social reproduction, such as healthcare, education,
pensions, and social security, deepening inequality in access to these
services (\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
\citeproc{ref-gingrich_making_2011}{Gingrich, 2011}). Yet public support
for market-based welfare provision has grown, even in traditional
welfare states, particularly among higher-income groups who view private
alternatives as more efficient or higher in quality
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).
This shift calls for broader inquiry into how institutions like markets
structure access to resources and legitimize inequalities
(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
\citeproc{ref-satz_por_2019}{Satz, 2019}).

The legitimacy of market-based inequalities is closely tied to beliefs
about distributive justice grounded in market principles. Beyond
stratification structures and socioeconomic conditions, the study of
inequality also focuses on individuals' beliefs about the origins of
inequality, the normative frameworks that sustain those beliefs, the
mechanisms that shape them, and their implications for attitudes and
behavior (\citeproc{ref-kluegel_beliefs_1981}{Kluegel \& Smith, 1981}).
In this regard, empirical research on distributive justice focuses on
individuals' conceptions of how goods and rewards \emph{should} be
distributed in society
(\citeproc{ref-jasso_distributive_2016}{Guillermina Jasso et al.,
2016}). This line of inquiry allows for examining the extent to which
economic inequality is perceived as just, or in a certain sense,
legitimate (\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011}).
Most of the research on this field explores the legitimacy of wage
inequality, specially salary gaps between jobs at opposite ends of the
occupational hierarchy
(\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011};
\citeproc{ref-jasso_justice_1978}{Guillermina Jasso, 1978};
\citeproc{ref-jasso_gender_1999}{G. Jasso \& Wegener, 1999}). Recently,
others examines how individuals justify market-generated inequalities in
access to core social services such as healthcare, education, pensions
or social security. Here, legitimacy stems from the belief that access
to these goods should follow market-based criteria
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such views, these
services are treated as legitimate commodities; goods that can be
traded, priced, and evaluated through market logic
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).

@@

The empirical study of market justice preferences has employed various
strategies to capture how individuals assess inequalities arising from
market allocation. A common approach evaluates whether people find it
fair that access to essential social services---such as healthcare,
education, or pensions---depends on income. This builds on foundational
work by Kluegel and Smith
(\citeproc{ref-kluegel_legitimation_1999}{1999}), who examined normative
justifications for capitalist systems. Subsequent studies have extended
this logic to a wider set of welfare goods. For example, Immergut and
Schneider (\citeproc{ref-immergut_it_2020}{2020}) and von dem Knesebeck
et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016}) explore whether
respondents believe it is just that higher-income individuals receive
better healthcare. Similarly, Lee and Stacey
(\citeproc{ref-lee_fairness_2023}{2023}) apply this framework to
education. Complementing these efforts, more recent research has
introduced composite indicators to capture broader orientations toward
market-based allocation. Lindh (\citeproc{ref-lindh_public_2015}{2015}),
for instance, constructs an index averaging support for income-based
access to healthcare and education in comparative perspective, while
Castillo et al. (\citeproc{ref-castillo_socialization_2024}{2024})
propose a single-item measure covering health, education, and pensions
in Chile. These instruments seek to gauge to what extent individuals
view market-generated inequalities as legitimate. Taken together, they
capture two core dimensions of market distribution: the role of economic
resources as a key determinant of outcomes, and the framing of services
as tradable commodities that can be bought and sold according to ability
to pay (\citeproc{ref-lindh_public_2015}{2015}).

@@

Research on intergenerational social mobility has increasingly examined
its effects on attitudes toward economic inequality. A key area of
inquiry focuses on how upward and downward mobility influence support
for redistribution, though findings remain mixed. Alesina et al.
(\citeproc{ref-alesina_intergenerational_2018}{2018}) show that
individuals with pessimistic expectations about their
mobility---particularly those anticipating downward movement---are more
likely to support generous redistributive policies. Similarly, Ares
(\citeproc{ref-ares_changing_2020}{2020}) finds that upwardly mobile
individuals tend to be less supportive of state-led redistribution
compared to those who have experienced downward mobility. Comparative
studies by Schmidt (\citeproc{ref-schmidt_experience_2011}{2011}) and
Gugushvili (\citeproc{ref-gugushvili_subjective_2017}{2017}) likewise
report that subjective upward mobility is associated with weaker
preferences for redistribution, while downward mobility strengthens
redistributive support. However, recent causal evidence from Breen and
Ermisch (\citeproc{ref-breen_effects_2024}{2024}) suggests the opposite:
upward mobility may increase redistributive preferences, while downward
mobility may reduce them.

@@

@@

\bibitem[\citeproctext]{ref-satz_por_2019}
Satz, D. (2019). \emph{{Por Qu{é} Algunas Cosas No Deber{í}an Estar en
Venta: Los l{í}mites Morales Del Mercado}}. Ciudad Aut{ó}noma de Buenos
Aires: Siglo XXI Editores.
